%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	ABSTRACT PAGE
%   TICKET-23 --- Abstract (~250 words)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\addtotoc{Abstract}
\abstract{\addtocontents{toc}{\vspace{1em}}

The Traveling Salesman Problem (TSP) is a classical combinatorial
optimisation problem that seeks the shortest closed tour visiting each
city exactly once.  Because the search space grows factorially with the
number of cities, exact methods become impractical for non-trivial
instances, making metaheuristic approaches such as Genetic Algorithms
(GAs) attractive alternatives.  A persistent challenge in applying GAs
to the TSP is that naive crossover operators produce infeasible
solutions --- chromosomes visiting some cities multiple times while
omitting others --- leading to meaningless fitness evaluations and
distorted selection pressure.

This report investigates the design, implementation, and critical
analysis of a GA for the static single-objective TSP with a focus on
feasibility repair mechanisms.  The primary GA pipeline pairs naive
single-point crossover with a random-insertion repair step that
diagnoses constraint violations (duplicates and missing cities) and
restores feasibility, so repair is invoked on every offspring.  We
compare this against the two alternative constraint-handling approaches
identified in the assignment specification: a penalty function that
inflates the fitness of infeasible chromosomes in proportion to
violations, and the feasibility-preserving Partially Mapped Crossover
(PMX) operator that guarantees valid offspring by construction.  Using
the kroA100 benchmark (100 cities, known optimal tour length
21\,282), we run a three-arm comparison of 30 independent seeds per
strategy alongside a 1\,620-run parameter sensitivity sweep over
population size, crossover rate, mutation rate, and selection method.
Repair achieves the lowest mean tour length ($49\,778$), with PMX close
behind ($50\,655$) and penalty noticeably worse ($57\,948$); all three
arms produce valid tours, so the differences reflect genuine quality
rather than infeasibility artefacts.  The findings show that direct
feasibility enforcement via repair can match or modestly outperform
feasibility-preserving operators on this benchmark, and substantially
outperform a penalty-based alternative.
}
\clearpage
