%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	CONCLUSION AND FUTURE RESEARCH DIRECTIONS
%   Mandated by Project Task 4.2 ("future research directions").
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lhead{Conclusion and Future Work}
\section{Conclusion and Future Research Directions}
\label{SecConclusion}

\subsection{Summary of Findings}
\label{Conc:Summary}

This project investigated a Genetic Algorithm for the static single-objective
Traveling Salesman Problem on the kroA100 benchmark instance (100 cities,
known optimal tour length 21\,282).  The primary pipeline pairs naive
single-point crossover with a random-insertion repair step that diagnoses
duplicates and missing cities and restores feasibility on every offspring.
Two alternative constraint-handling strategies were implemented as
comparators: a penalty function that inflates the fitness of infeasible
chromosomes, and the feasibility-preserving Partially Mapped Crossover
operator (PMX).

The main findings are as follows:

\begin{enumerate}[nosep]
    \item \textbf{Repair achieves the lowest tour length among the three
    constraint-handling strategies.}
    In the three-arm comparison (\sref{T4:Boxplots}), the repair arm
    produced a mean best fitness of $49\,778$, with PMX close behind at
    $50\,655$ and the penalty arm noticeably worse at $57\,948$.  All
    three arms returned valid tours at the end of every run, so the gaps
    reflect genuine differences in search quality rather than infeasibility
    artefacts.

    \item \textbf{The assignment's repair-centric pipeline is competitive
    with the architecturally clean PMX alternative.}
    The gap between repair ($49\,778$) and PMX ($50\,655$) is small
    relative to the gap to penalty ($57\,948$).  Direct feasibility
    enforcement is therefore a credible alternative to a feasibility-
    preserving operator, while penalty-based handling lags both.

    \item \textbf{Selection method is the dominant parameter.}
    The $1\,620$-run sensitivity sweep (\sref{T3:Sensitivity}) shows
    tournament selection strongly outperforming roulette across all
    population sizes, crossover rates, and mutation rates.

    \item \textbf{Mutation rate is the second-order parameter; population
    size third.}
    Higher mutation rates reduce premature convergence and improve mean
    tour length, while population size has a smaller but consistent
    monotonic effect.  Crossover rate had the weakest standalone effect
    over the swept range.

    \item \textbf{Best configuration.}
    Population size $200$, crossover rate $0.95$, mutation rate $0.10$,
    tournament selection achieved a mean best fitness of $49\,378$ across
    seeds (mean gap $131.9\%$ from optimal), with the best single seed
    reaching $42\,623$ (gap $100.3\%$).
\end{enumerate}

The headline conclusion is that the assignment's repair-centric primary
pipeline is competitive with feasibility-preserving PMX on this benchmark
and substantially better than penalty-based handling.

\subsection{Limitations}
\label{Conc:Limitations}

Several limitations should be noted when interpreting these results.

\textbf{Single benchmark instance.}
All experiments were conducted on kroA100.  The findings may not generalise
to larger TSPLIB instances, asymmetric TSP variants, or instances with
substantially different city distributions.

\textbf{Single repair strategy in the production pipeline.}
Three repair insertion strategies were implemented (random, positional,
nearest-neighbour), but only random insertion was used in the experiments
reported here.  The remaining two were not exercised in the production
pipeline.

\textbf{No local-search hybridisation.}
The best mean gap of $\sim\!100\%$ from optimal reflects the absence of
2-opt or Lin--Kernighan refinement rather than a flaw in the repair
mechanism itself.  A plain generational GA, with or without repair, is not
expected to close this gap on a 100-city instance.

\textbf{Diversity tracked only via Hamming distance.}
Population diversity was monitored using pairwise Hamming distance over
chromosome positions.  \cite{cicirello2023permutationoperators} argues
that edge-based diversity is more TSP-specific because fitness depends on
adjacent city pairs rather than absolute positions; this project does not
implement edge-based metrics.

\textbf{Naive crossover as primary by design.}
Naive single-point crossover was deliberately chosen as the primary
operator so that the repair mechanism is load-bearing rather than
decorative.  A natural alternative --- using PMX or Order Crossover (OX) as
the primary operator with repair as a safety net --- is implemented but is
not the configuration evaluated in the three-arm comparison.

\textbf{Penalty arm uses a single fixed weight.}
The penalty function applies a fixed multiplicative weight of $50\,000$
per violation.  \cite{chehouri2016violationfactor} reports that adaptive
penalty schemes that scale with constraint violation magnitude can
outperform fixed weights; this was not tested.

\subsection{Future Research Directions}
\label{Conc:FutureWork}

The following directions extend the current work:

\begin{enumerate}[nosep]
    \item \textbf{Edge-based diversity instrumentation.}
    Add edge-overlap and edge-diversity metrics alongside Hamming, and
    re-run the three-arm comparison and sensitivity sweep under the new
    instrumentation.  TSP fitness depends on the set of edges in a tour,
    so edge-based metrics should track convergence more faithfully than
    position-based ones (\cite{cicirello2023permutationoperators}).

    \item \textbf{Hybrid local search.}
    Integrate 2-opt or Lin--Kernighan refinement after crossover and
    repair, applied either to every offspring or to elite individuals
    only.  This targets the $\sim\!100\%$ optimality gap directly and
    is the standard route to competitive TSP results in the GA literature
    (\cite{liu2014powerfulga}).

    \item \textbf{Adaptive repair strategy.}
    Empirically compare random, positional, and nearest-neighbour
    insertion across the same seeds and configurations used here, then
    design an adaptive variant that selects an insertion rule based on
    current population diversity or on the recent fitness contribution
    of each rule.

    \item \textbf{Larger and structurally different benchmarks.}
    Extend the experimental harness to kroA200, kroA500, and an
    asymmetric TSP instance.  This would reveal whether the observed
    ranking (repair $\succ$ PMX $\succ$ penalty) holds at larger scales
    and under asymmetric distance matrices.

    \item \textbf{Adaptive penalty weights.}
    Implement the violation-factor approach from
    \cite{chehouri2016violationfactor} and compare it against the
    fixed-weight baseline used in the penalty arm.  This isolates whether
    the penalty arm's poor showing reflects an intrinsic limitation of
    penalty-based handling or simply an untuned weight.

    \item \textbf{Operator interaction studies.}
    The sensitivity sweep fixed \texttt{crossover\_method} at naive and
    varied population size, crossover rate, mutation rate, and selection
    method.  A crossed sweep that also varies the crossover operator
    (naive, PMX, OX, CX) would quantify operator-by-parameter
    interactions directly, and would clarify whether the parameter
    rankings reported here hold under feasibility-preserving operators.
\end{enumerate}
