%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	LITERATURE REVIEW
%   Mandated by Project Task 4.2 ("literature review on Genetic Algorithms and
%   constraint handling techniques").
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lhead{Literature Review}
\section{Literature Review}
\label{SecLiteratureReview}

The Traveling Salesman Problem (TSP) has long been used as a
benchmark for combinatorial optimisation because it combines a simple
objective function with a large permutation search space. Exact approaches
can solve selected benchmark instances, but the factorial growth of the
search space motivates metaheuristic methods for repeated experimentation
and algorithm analysis~\cite{applegate2006traveling,reinelt1991tsplib}.
Genetic Algorithms (GAs), introduced by Holland and popularised by
Goldberg, are particularly relevant because they search with a population of
candidate tours rather than a single incumbent solution
\cite{holland1975adaptation,goldberg1989genetic}. For the TSP, however,
the success of a GA depends strongly on how tours are represented, how
crossover and mutation preserve permutation feasibility, and how constraint
violations are handled when infeasible offspring are created.

\subsection{Genetic Algorithms for the TSP}
\label{Lit:GAforTSP}

The most common GA representation for the TSP is a permutation of city
indices. This direct encoding is natural because each chromosome position
contains one city and a complete chromosome defines the visit order.
However, this representation also makes the problem sensitive to genetic
operator design: a chromosome is feasible only when every city appears
exactly once. The classic review by Larra{\~n}aga et al. shows that GA
research for the TSP developed around this representation issue, comparing
path, adjacency, ordinal, matrix, and other encodings together with the
operators that act on them~\cite{larranaga1999genetic}. Later reviews and
historical summaries similarly identify crossover design, mutation design,
hybrid local search, and diversity preservation as recurring themes in
GA-TSP work~\cite{scholz2019historicalreview,cicirello2023permutationoperators}.

Permutation-specific recombination is important because ordinary string
crossover can destroy the city set. Early operators such as Partially Mapped
Crossover (PMX), Order Crossover (OX), and Cycle Crossover (CX) were
therefore developed to recombine parent tours while retaining a valid
permutation~\cite{goldberg1985alleles,davis1985applying,oliver1987study}.
These operators preserve different structures. PMX emphasizes positional
mapping, OX preserves relative order, and CX preserves cycles of
position-wise inheritance. More recent operator surveys make this distinction
explicit by classifying permutation operators according to the features they
transmit, including absolute positions, undirected edges, directed edges,
precedences, and cyclic precedences~\cite{cicirello2023permutationoperators}.
This is directly relevant to the present project because TSP tour length is
computed from adjacent city pairs. Positional diversity captured by Hamming
distance describes how the population spreads across the city-position
space, which is the diversity signal this project tracks across runs.

State-of-the-art GA variants for the TSP often go beyond a plain
generational GA. Liu uses edge swapping and local search to obtain strong
results on large TSP benchmarks, emphasising that edge-based structure is
central to tour quality~\cite{liu2014powerfulga}. Eremeev and Kovalenko
similarly show, for asymmetric TSP instances, that recombination design,
neighbourhood-based mutation, local search, restarts, and population
diversity all influence performance~\cite{eremeev2017optimalrecombination}.
These studies set context rather than direct performance targets for this
coursework project: the implementation here remains intentionally focused on
GA fundamentals, repair mechanisms, and comparative constraint handling on
kroA100.

\subsection{Constraint Handling in Evolutionary Algorithms}
\label{Lit:ConstraintHandling}

Constraint handling is a general challenge in evolutionary computation.
Coello Coello surveys theoretical and numerical techniques used with
evolutionary algorithms, including penalty functions, special
representations, repair methods, and feasibility-based
selection~\cite{coello2002constraint}. Michalewicz and Schoenauer similarly
argue that constraints are not an implementation detail but part of the
problem structure that must be reflected in representation, operators, or
evaluation~\cite{michalewicz1996evolutionary}. These ideas apply directly
to the TSP: the ``visit every city exactly once'' condition is not a minor
side constraint, but the property that makes a chromosome represent a valid
tour.

Penalty functions are one common approach. They leave infeasible solutions
in the population but add a cost proportional to constraint violation. This
approach is easy to implement, but it introduces a tuning problem: the
penalty must be large enough to discourage infeasible solutions but not so
large that the search loses useful information too early. Chehouri et al.
make this critique explicit and propose a violation-based alternative to
avoid penalty-parameter tuning in constrained GA
optimisation~\cite{chehouri2016violationfactor}. In the TSP setting, this
trade-off is especially visible. If a chromosome skips cities because of
duplicates, its raw tour length may be artificially low, so weak penalties
can reward invalid tours. Excessive penalties, however, may remove
offspring that contain useful partial order or edge information.

Another approach is to prevent infeasibility by construction. PMX, OX, CX,
and other permutation-aware operators are examples of feasibility-preserving
operators for TSP chromosomes. This approach is attractive because every
offspring can be evaluated directly as a valid tour. Its limitation is that
it restricts the set of admissible operators and may make it harder to study
the effect of infeasible offspring. For this reason, the present project
uses both feasibility-preserving operators and a deliberately naive crossover
baseline, so that the effect of repair can be observed rather than assumed.

\subsection{Repair Mechanisms in Permutation-based EAs}
\label{Lit:Repair}

Repair mechanisms form a third constraint-handling strategy. Instead of
discarding or penalising infeasible chromosomes, repair modifies them into
feasible candidates before evaluation. Padhye et al. discuss
feasibility-preserving repair strategies in evolutionary optimisation and
highlight an important caveat: repair can restore feasibility, but it can
also introduce search bias or reduce exploration if repaired solutions are
systematically pushed into particular regions of the feasible
space~\cite{padhye2015feasibilitypreserving}. Salcedo-Sanz's survey of
repair methods makes the same issue central by treating repair as a broad
class of constraint-handling techniques rather than a simple preprocessing
step~\cite{salcedosanz2009repair}.

For permutation problems, repair usually acts on duplicate, missing, or
misordered genes. In the TSP case studied here, a naive crossover can create
offspring with duplicated cities and omitted cities. A repair mechanism can
scan the chromosome, retain one copy of each duplicated city, and insert the
missing cities into the remaining positions. This restores the formal TSP
constraint and prevents invalid short tours from dominating selection.
However, repair is not neutral. Random insertion may preserve diversity but
disturb useful edges. Positional insertion may retain more chromosome
structure but reinforce parent-position bias. Nearest-neighbour insertion
may improve local tour quality but also accelerate convergence toward a
narrower set of edges. These trade-offs motivate the project's explicit
tracking of best fitness, mean fitness, and Hamming diversity across runs.

The critical point for this project is that feasibility and quality are
different properties. A repaired chromosome is valid, but it is not
necessarily a high-quality tour, and the repair operation may alter the
building blocks inherited from the parents. Recent TSP crossover work
reinforces this distinction by showing that tour symmetries, rotations,
reversals, and edge structure affect whether fit parents produce fit
offspring~\cite{uray2023csrx}. Thus, repair should be evaluated not only by
whether it produces feasible tours, but also by its effect on solution
quality, diversity, and convergence.

\subsection{Comparison Studies of TSP Operators}
\label{Lit:CrossoverStudies}

Empirical crossover and mutation studies provide practical context for the
operator choices in this project. Abdoun and Abouchabaka compare several
crossover operators for the TSP, including OX, PMX, UPMX, CX, and NWOX, and
report strong performance for OX on their benchmark
setting~\cite{abdoun2012adaptivecrossover}. Abdoun, Abouchabaka, and Tajani
perform a related mutation-operator study, emphasising that representation,
selection, crossover, mutation, and operator probabilities jointly define a
GA variant~\cite{abdoun2012mutationoperators}. These findings support the
project's modular notebook structure, where representation, selection,
crossover, mutation, diversity metrics, repair, and experiments are separated
so that their effects can be analysed.

More recent crossover studies also use TSPLIB-scale instances similar to
this project. Thanh et al. test combined crossover mechanisms on Euclidean
TSPLIB instances ranging from 51 to 575 cities, including kroA100 and
lin105~\cite{thanh2020combinationcrossover}. Their results reinforce that
operator design can affect both mean and best tour quality. Uray et al.
extend this line of work by designing crossover operators around symmetric
TSP invariances such as circular shifts and reversal of tours
\cite{uray2023csrx}. Together, these studies justify treating crossover as a
core experimental factor rather than a fixed implementation detail.

In summary, the literature supports the design of this project in three
ways. First, GA-TSP performance depends on permutation-aware operators and
population diversity. Second, infeasible chromosomes can distort selection
pressure if evaluated directly. Third, repair mechanisms are useful but must
be analysed critically because restoring feasibility can change diversity,
edge structure, and convergence behaviour. These points motivate the
repair-focused experiments and the comparative analysis reported in the
following sections.
