%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	TASK 2 - Algorithm Development
%   Maps to brief Tasks 2.1 (GA) and 2.2 (Constraint Handling via Repair).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lhead{Task 2. \emph{Algorithm Development}}
\section{Task 2: Algorithm Development}
\label{Task2}

This chapter specifies the genetic algorithm used throughout the project and
the repair mechanism that handles the permutation constraint. Section
\ref{T2:GADesign} fixes the representation and the five evolutionary
operators. Section \ref{T2:Repair} explains why the primary recombination
operator produces infeasible offspring on every generation and describes the
repair step that converts those offspring back into valid Hamiltonian tours.
The empirical comparison of repair against alternative constraint-handling
approaches is reported in Section \ref{T3:RepairCritique}.

\subsection{Genetic Algorithm Design (Task 2.1)}
\label{T2:GADesign}

\subsubsection{Solution Representation}
\label{T2:Representation}
Each candidate solution is encoded as a permutation of the city index set
$\{0, 1, \ldots, n-1\}$. A chromosome is an ordered tuple in which the
$i$-th gene names the $i$-th city visited; the tour is closed by returning
to the first gene. For the eight-city illustration used later, the
chromosome $[3,0,7,1,5,2,6,4]$ denotes the tour
$3 \to 0 \to 7 \to 1 \to 5 \to 2 \to 6 \to 4 \to 3$. The fitness of a
chromosome $\pi$ is its closed Euclidean tour length,
\begin{equation}
f(\pi) \;=\; \sum_{i=0}^{n-1} d_{\pi_i,\,\pi_{(i+1) \bmod n}},
\end{equation}
where $d_{ij}$ is the Euclidean distance between cities $i$ and $j$.
Because fitness is computed by indexing the distance matrix at the gene
values, the calculation is only meaningful when the chromosome is a true
permutation. A chromosome with duplicate or missing cities still produces a
number, but that number is not the length of a Hamiltonian tour and must
not be admitted into selection (see Section \ref{T2:Infeasibility}).

\subsubsection{Population Initialisation}
\label{T2:Init}
The initial population is generated as $N$ independent random permutations
of the city set, produced with
\texttt{numpy.random.default\_rng(seed).permutation(n)}. Each chromosome is
sampled independently, so the initial population is drawn uniformly from
the space of $n!$ permutations and starts at near-maximal diversity. No
heuristic seeding (such as a nearest-neighbour tour) is injected, which
keeps the convergence behaviour reported in Section \ref{T4:Convergence}
attributable to the evolutionary operators rather than to a warm-started
population. A single seed controls every stochastic component of the GA,
making each run reproducible and supporting paired-seed comparisons in
Section \ref{T3:Sensitivity}.

\subsubsection{Selection}
\label{T2:Selection}
Two selection mechanisms are implemented and exposed through a single
configuration parameter, so that the choice of parent-selection rule can be
swept in the same way as any numerical hyperparameter.

Tournament selection is the primary operator. Given a population $P$ of
size $N$, $k$ candidates are drawn uniformly at random without replacement
and the candidate with the lowest tour length is returned,
\begin{equation}
s \;=\; \arg\min_{i \in C} f(i), \qquad C \subseteq P,\; |C| = k.
\end{equation}
The tournament size $k$ controls selection pressure; small $k$ keeps weak
individuals in play, large $k$ narrows the gene pool quickly. The primary
configuration uses $k = 3$, which retains enough randomness to delay
premature convergence while still favouring the better half of the
population. The sweep in Section \ref{T3:Sensitivity} varies $k$
explicitly.

Roulette wheel selection is implemented as a comparison operator. Because
the TSP is a minimisation problem, raw fitness cannot be used as a
selection weight directly; instead each individual is assigned the inverse
fitness and the weights are normalised to a probability distribution,
\begin{equation}
p_i \;=\; \frac{1/f_i}{\sum_{j=1}^{N} 1/f_j}.
\end{equation}
The roulette draw then picks a parent with probability $p_i$. The same
unified \texttt{select} function dispatches on the
\texttt{selection\_method} field of the GA configuration, so the loop
body is identical for both methods.

\subsubsection{Crossover}
\label{T2:Crossover}
The primary recombination operator is naive single-point crossover.
Given two parents $A$ and $B$ of length $n$ and a cut point
$pt \in \{1, \ldots, n-1\}$, the child is constructed by concatenation,
\begin{equation}
\mathrm{child} \;=\; A[\,0\,{:}\,pt\,] \,\Vert\, B[\,pt\,{:}\,n\,].
\end{equation}
This operator copies positions independently, without examining whether the
inserted gene values have already appeared elsewhere in the child. It does
not preserve the permutation property: the same city may appear twice and
another may disappear. The choice is deliberate. Pairing naive crossover
with the repair step makes the repair mechanism load-bearing on every
generation rather than a decorative safety net invoked once in a thousand
matings, which matches the assignment's focus on repair as the central
constraint-handling contribution.

Three feasibility-preserving operators are also implemented as
alternatives for the comparison in Section \ref{T3:RepairCritique}.
Partially Mapped Crossover (PMX) of \cite{goldberg1985alleles} copies a
segment $A[p_1{:}p_2]$ into the child and routes each displaced value from
$B$ through the segment mapping until it reaches an unoccupied position;
remaining slots are filled directly from $B$. Order Crossover (OX), due to
\cite{davis1985applying} and refined in \cite{oliver1987study}, also copies
the segment from $A$ and then walks $B$ in wrap-around order from just
after $p_2$, skipping any value already present in the segment. Cycle
Crossover (CX) discovers position-wise cycles between the two parents and
inherits whole cycles alternately from each parent, so every position is
filled once and every city appears once. The constructive arguments behind
PMX, OX, and CX, together with comparative reviews of permutation
operators, are summarised in \cite{larranaga1999genetic} and
\cite{cicirello2023permutationoperators}; the worked example in Section
\ref{T2:Infeasibility} shows the three operators producing valid
permutations on the same parents that defeat naive crossover.

\subsubsection{Mutation}
\label{T2:Mutation}
Swap mutation is the sole mutation operator used in the primary pipeline.
On each invocation a uniform variate is drawn from $[0,1)$; if it falls
below the mutation rate $r_m$, two distinct gene positions $i$ and $j$ are
sampled without replacement and the gene values at those positions are
exchanged. The chromosome length, the set of cities, and the permutation
property are all preserved by construction, so swap mutation introduces no
infeasibility of its own. Its perturbation is local: exactly two edges in
the tour change, which makes it well suited to late-stage refinement.
Inversion mutation, in which the segment between two cut points is
reversed in place, is implemented in the code base as a comparison
operator but is not used by any of the primary experiments in this report.

\subsubsection{Main Loop}
The five operators are combined into a generational GA with elitism.
At each generation the top $e$ individuals are copied unchanged into the
next population; the remainder is generated by repeated selection,
crossover (with probability $r_c$), mutation, and (if enabled) repair.
The elitism count is fixed at $e = 2$ throughout the primary experiments,
which protects the incumbent best while leaving most of the population
open to recombination. Algorithm \ref{alg:ga} states the loop precisely.

\begin{algorithm}[h]
\caption{Generational GA with Optional Repair}
\label{alg:ga}
\begin{algorithmic}[1]
\State Initialise population $P$ of $N$ random permutations
\State Evaluate fitness $f(p)$ for all $p \in P$
\For{$g = 1$ to $G_{\max}$}
    \State $P' \gets$ top-$e$ elites from $P$ \Comment{elitism}
    \While{$|P'| < N$}
        \State Select parents $p_1, p_2$ from $P$ via tournament or roulette
        \If{$\mathrm{rand}() < r_c$}
            \State $c \gets \mathrm{Crossover}(p_1, p_2)$ \Comment{naive single-point in the primary pipeline}
        \Else
            \State $c \gets \mathrm{copy}(p_1)$
        \EndIf
        \State $c \gets \mathrm{Mutate}(c, r_m)$ \Comment{swap mutation}
        \If{repair enabled}
            \State $c \gets \mathrm{Repair}(c)$
        \EndIf
        \State Append $c$ to $P'$
    \EndWhile
    \State $P \gets P'$; evaluate fitness for all $p \in P$
\EndFor
\State \Return best individual in $P$
\end{algorithmic}
\end{algorithm}

\subsection{Constraint Handling via Repair Mechanism (Task 2.2)}
\label{T2:Repair}

The TSP imposes a single hard constraint on the chromosome: every city
must appear exactly once. The constraint cannot be relaxed without
changing the problem, so the GA either uses recombination operators that
preserve it by construction or repairs violations after they occur.
Surveys of evolutionary constraint handling describe the same trichotomy
of options --- preservation, repair, and penalty
\cite{coello2002constraint,michalewicz1996evolutionary}. The primary
pipeline of this project takes the repair route, paired with naive
single-point crossover so that the repair step does real work on every
offspring.

\subsubsection{Why Infeasibility Arises}
\label{T2:Infeasibility}
Naive single-point crossover treats the chromosome like an ordinary
fixed-position string. It copies gene positions independently and does
not check whether the inserted value already appears elsewhere in the
child. For a permutation representation the constraint is global rather
than positional, so this position-by-position copy is exactly the
operation that breaks the city set.

Consider the following worked example. The parents are
\begin{align*}
A &= (0,1,2,3,4,5,6,7) \\
B &= (2,6,4,0,5,7,1,3)
\end{align*}

and the single cut point is $pt = 4$. The child constructed by naive
single-point crossover is

\begin{align*}
\mathrm{child} &= A[\,0{:}4\,] \,\Vert\, B[\,4{:}8\,] \\
&= (0,1,2,3,5,7,1,3)
\end{align*}

City $1$ now appears at positions $1$ and $6$, and city $3$ appears at
positions $3$ and $7$. Cities $4$ and $6$ never appear. The chromosome
has the correct length but is not a Hamiltonian tour. The same
mechanism causes violations on kroA100, where every one
of thirty naive-crossover offspring was infeasible, with twenty-four
duplicates on average.

The downstream effect on selection is the problem. Fitness evaluation
on $(0,1,2,3,5,7,1,3)$ still returns a number, but the number sums
distances over an edge set that visits city $1$ twice, visits city $3$
twice, and skips cities $4$ and $6$ entirely. The result tends to be
artificially small: cheap edges contribute repeatedly while expensive
missing-city edges are dropped. Selection compares this number against
the genuine tour lengths of feasible competitors as if they were
commensurable. Tournament and roulette selection then favour the
infeasible chromosome, its invalid genetic material spreads through
recombination, and within a handful of generations the population
converges towards short pseudo-tours rather than short tours. The
infeasibility analysis in
\cite{cicirello2023permutationoperators} traces the same failure mode
across the permutation operator family.

The contrast with the permutation-preserving operators makes the
mechanism clear. On the same parents and the same cut points: 

\begin{itemize}
	\item PMX returns $(5,6,2,3,4,7,1,0)$.
	\item OX returns $(0,5,2,3,4,7,1,6)$.
	\item CX returns $(0,6,2,3,4,5,1,7)$
\end{itemize}


All three are valid permutations because
each operator embeds a feasibility-preserving rule into the copy step
itself. The repair mechanism described below recovers the same
guarantee by acting after a naive copy rather than during one.

\subsubsection{Repair Design}
\label{T2:RepairDesign}
The repair function follows a two-phase structure: diagnose the
infeasible chromosome, then correct it. The diagnosis phase is a single
left-to-right scan. A \emph{seen} set records the first occurrence of
each city in scan order; any subsequent occurrence of the same city is
flagged as a duplicate position. After the scan, the missing set is
computed as $\{0, 1, \ldots, n-1\} \setminus \mathrm{seen}$. Because
the chromosome length is fixed at $n$, the number of duplicate positions
equals the number of missing cities, so every duplicate can be replaced
with a unique missing city.

The correction phase assigns the missing cities to the duplicate
positions. The order in which the assignment is made distinguishes
three strategies:
\begin{enumerate}[nosep]
    \item \textbf{Random insertion} (the primary strategy) shuffles the
          missing cities with the GA's seeded random generator and
          places them at the duplicate positions in shuffled order.
    \item \textbf{Position insertion} places the missing cities in
          sorted order at the duplicate positions in left-to-right
          order. The assignment is deterministic, so the same
          infeasible chromosome always repairs to the same tour.
    \item \textbf{Nearest-neighbour insertion} processes the duplicate
          positions left to right and, at each position, chooses the
          still-unassigned missing city whose Euclidean distance to the
          immediately preceding gene in the chromosome is smallest.
\end{enumerate}
Random insertion is the primary choice. It introduces no bias from the
problem geometry, so the repair step contributes feasibility without
contributing optimisation; the GA's selection pressure remains the sole
driver of solution quality. Random insertion also tends to preserve
population diversity better than the deterministic alternatives,
because two infeasible chromosomes with the same duplicates and missing
cities can still be repaired into different tours. The
feasibility-preserving repair literature reports the same tension
between feasibility and diversity
\cite{padhye2015feasibilitypreserving}. Algorithm \ref{alg:ga} places
the repair call after mutation, so any chromosome reaching the fitness
evaluation step at the end of the loop is guaranteed to be a valid
permutation.

The repair function is illustrated step by step below. On the eight-city chromosome
$(3,1,4,1,5,2,6,3)$, the diagnosis phase identifies duplicate positions
$\{3, 7\}$ and missing cities $\{0, 7\}$. Random insertion may return
$(3,1,4,7,5,2,6,0)$; position insertion deterministically returns
$(3,1,4,0,5,2,6,7)$; nearest-neighbour insertion on the toy octagon
returns $(3,1,4,7,5,2,6,0)$. All three are valid permutations. A
property test runs the repair on five hundred random
corruptions across all three strategies and observes a valid
permutation in every case.

\subsubsection{Effect on Quality, Diversity, and Convergence}
\label{T2:RepairEffect}
The repair step has three intended effects on the GA's behaviour. The
quantitative measurement of those effects is reported in Sections
\ref{T3:RepairCritique} and \ref{T4:Convergence}; this section sets
expectations.

The first effect is on fitness validity. Every chromosome that reaches
the fitness evaluation step in Algorithm \ref{alg:ga} is a true
permutation, so the tour-length number returned by the distance matrix
indexing is the actual length of a Hamiltonian tour over $n$ cities.
Without repair, the same fitness function returns a number whose
relationship to a tour length is corrupted by the same duplicate and
missing-city counting that produced it; the apparent tour lengths are
smaller because they sum over fewer edges, not because the underlying
tours are better. Section \ref{T3:RepairCritique} confirms this
empirically: the no-repair arm posts a lower mean fitness than the
repair arm at generation five hundred, but the no-repair fitness does
not correspond to a valid 100-city tour and is not comparable.

The second effect is on diversity. Random insertion injects stochastic
variation at the duplicate positions, so a population that would
otherwise lose diversity through repeated crossovers between similar
chromosomes retains some Hamming distance. The measured diversity
trajectory in Section \ref{T3:RepairCritique} shows the repair arm
holding slightly higher Hamming distance than the no-repair baseline at
the same generation, despite both arms losing most of the initial
diversity by generation five hundred. Deterministic repair strategies
do not provide this benefit, which is part of the reason random
insertion is the primary choice rather than position insertion.

The third effect is on convergence. The repair step keeps the search
inside the feasible region, so the apparent improvement in best fitness
across generations reflects actual tour-length reduction rather than an
increasing severity of constraint violation. In the comparison reported
in Section \ref{T3:RepairCritique}, the repair pipeline converges to
solutions comparable with the feasibility-preserving PMX alternative
and narrowly outperforms it on the kroA100 paired-seed runs. The
penalty function comparator, in contrast, exhibits the failure mode
discussed in
\cite{coello2002constraint,michalewicz1996evolutionary}: the penalty
weight is hard to tune, and the GA converges either to overpenalised
feasible-but-poor regions or to underpenalised infeasible regions.
Section \ref{T3:RepairCritique} reports the magnitudes; this section
records that repair was retained as the primary mechanism on the basis
of feasibility validity, diversity preservation, and stable
convergence.
