%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	TASK 1 - Problem Definition and Experimental Design
%   Maps to brief Tasks 1.1, 1.2, 1.3.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lhead{Task 1. \emph{Problem Definition and Experimental Design}}
\section{Task 1: Problem Definition and Experimental Design}
\label{Task1}

\subsection{Benchmark TSP Instance (Task 1.1)}
\label{T1:Benchmark}

The benchmark instance used throughout this project is \textbf{kroA100}, drawn
from the TSPLIB collection compiled by Reinelt~\cite{reinelt1991tsplib} and
described in the comprehensive treatment of Applegate et
al.~\cite{applegate2006traveling}. The instance contains 100 cities placed in
the Euclidean plane, uses the symmetric \texttt{EUC\_2D} edge-weight type, and
has a known optimal tour length of 21{,}282. Its dimension lies within the
70--150 range required by the project brief.

The choice of kroA100 is deliberate. It is one of the most frequently cited
TSPLIB instances in the GA-for-TSP literature, which means its random
Euclidean layout, its optimum, and its level of difficulty are well documented
and directly comparable to prior empirical studies. The size of $n=100$ is
large enough to make exhaustive enumeration impossible
($99!/2 \approx 4.7\times 10^{155}$ distinct tours), yet small enough to
support a 54-configuration parameter sweep with 30 independent runs each
within a practical compute budget. A known optimum allows results to be
reported as a relative gap, which is more interpretable than absolute tour
length in isolation.

\begin{table}[h]
	\centering
	\caption{kroA100 benchmark instance metadata.}
	\label{tab:instance}
	\begin{tabular}{ll}
		\hline
		\textbf{Property} & \textbf{Value} \\
		\hline
		Name & kroA100 \\
		Number of cities ($n$) & 100 \\
		Edge-weight type & \texttt{EUC\_2D} (symmetric Euclidean 2D) \\
		Known optimal tour length & 21{,}282 \\
		Instance class & Symmetric Euclidean TSP \\
		Source & TSPLIB~\cite{reinelt1991tsplib,applegate2006traveling} \\
		\hline
	\end{tabular}
\end{table}

For \texttt{EUC\_2D} instances, each city $i$ has coordinates $(x_i, y_i)$ and
the inter-city distance follows the TSPLIB convention of nearest-integer
rounding,
\[
d(i,j) = \left\lfloor \sqrt{(x_i - x_j)^2 + (y_i - y_j)^2} + 0.5 \right\rfloor,
\]
so that $d(i,j) = d(j,i)$. The same rounded distance is used for fitness
evaluation and for comparison against the published optimum.

\subsection{Formal Mathematical Formulation (Task 1.2)}
\label{T1:Formulation}

The symmetric TSP on kroA100 is formulated directly as a permutation
optimisation problem, which matches the GA representation used in the
remainder of the report. Let $V = \{0, 1, \ldots, n-1\}$ denote the set of
city indices with $n=100$, and let $d : V \times V \to \mathbb{Z}_{\geq 0}$
denote the rounded Euclidean distance defined above.

A candidate solution is a tour $\pi = (\pi_0, \pi_1, \ldots, \pi_{n-1})$,
where $\pi$ is a permutation of $V$. The fitness function to be minimised is
the total length of the closed tour induced by $\pi$,
\[
\min_{\pi} \; f(\pi)
\;=\;
\sum_{i=0}^{n-1} d\!\left(\pi_i,\, \pi_{(i+1) \bmod n}\right),
\]
where the modular index in the final term closes the tour by linking
$\pi_{n-1}$ back to $\pi_0$.

Two constraints are implicit in this formulation. The permutation constraint
requires that every city is visited exactly once, i.e.\
$\{\pi_0, \pi_1, \ldots, \pi_{n-1}\} = V$ with no repetitions. The
closed-tour constraint is captured by the modular indexing and forces the
salesman to return to the starting city. Together they replace the
incoming-degree, outgoing-degree, and subtour-elimination constraints of the
classical arc-based integer programming formulation: any permutation of $V$
is automatically a Hamiltonian cycle, so feasibility reduces to maintaining
the permutation property under the GA operators.

The properties of this search space drive every subsequent design choice.
The space of feasible solutions is discrete and finite, with cardinality
$n!/(2n)$ once rotational and reflective symmetries of the symmetric tour
are quotiented out; for $n=100$ this is approximately
$4.7 \times 10^{155}$. The factorial growth rules out exact search at this
scale and motivates a population-based metaheuristic such as a Genetic
Algorithm~\cite{holland1975adaptation,goldberg1989genetic}. The fitness
landscape is non-convex and multimodal: many tours share similar lengths but
differ structurally in the edges they use, which is the regime in which
recombination of partial tours can outperform a single-point local search.
The permutation constraint, finally, dictates the operator design. Classical
string-style crossover and bit-flip mutation do not preserve the permutation
property, so the algorithm must either use feasibility-preserving operators
or pair an unconstrained operator with an explicit repair step. The present
project takes the latter route in order to study repair as an explicit
experimental factor, following the constraint-handling taxonomy reviewed
by Larra{\~n}aga et al.~\cite{larranaga1999genetic}.

\subsection{Experimental-Setup Justification (Task 1.3)}
\label{T1:Setup}

The experimental protocol is fixed before any results are collected so that
every configuration is evaluated on identical terms.

Each configuration is run for \textbf{30 independent random seeds}. Thirty
runs is the smallest sample size at which the central limit theorem provides
a reasonable normal approximation for the sampled mean tour length, which
allows means, standard deviations, and confidence intervals to be reported
without leaning on distributional assumptions about the underlying tour-length
distribution. It is also small enough to keep the full sweep tractable: a
larger sample would buy diminishing returns in mean-estimate stability while
multiplying the compute cost linearly across all configurations.

Each run terminates at a fixed budget of \textbf{500 generations}.
Preliminary convergence curves, reported in §\ref{T4:Convergence}, show that
the bulk of fitness improvement occurs within roughly the first 200
generations across all configurations; the remaining 300 generations cover
the long tail in which late mutations or rare crossover events still trim a
small number of edges. Using a fixed generation budget rather than an
adaptive stopping rule ensures that every run produces the same number of
per-generation observations, which keeps convergence plots, diversity curves,
and statistical tests directly comparable across configurations.

The parameter sweep covers four GA factors at three or two levels each,
giving $3 \times 3 \times 3 \times 2 = 54$ unique configurations. The factor
levels are
\[
\text{pop\_size} \in \{50, 100, 200\}, \qquad
p_c \in \{0.70, 0.85, 0.95\},
\]
\[
p_m \in \{0.01, 0.05, 0.10\}, \qquad
\text{selection} \in \{\text{tournament},\, \text{roulette}\}.
\]
Population size brackets the moderate range typical in GA-TSP studies on
TSPLIB-scale instances. Crossover and mutation rates span the
low/medium/high settings used in the empirical reviews of Larra{\~n}aga et
al.~\cite{larranaga1999genetic}, which lets the sweep separate the effect of
exploration pressure from the effect of disruption. Selection method is
included at two levels because tournament and roulette selection apply
qualitatively different selection pressure profiles, and the interaction
between selection pressure and mutation rate is one of the questions Task 4
will analyse.

Combining the sweep with the seed replication gives
$54 \times 30 = 1{,}620$ total runs, which is large enough to support
non-parametric tests of configuration differences and a per-factor
sensitivity analysis.

A second group of parameters is held fixed across the sweep because the
project's design intent is to isolate the four factors above rather than
explore the full operator space. The tournament size is fixed at $k=3$, the
number of elite individuals carried over each generation is fixed at 2,
crossover uses the deliberately naive single-point operator with the repair
step enabled, and mutation uses the swap operator. The choice of a naive
single-point crossover with repair is itself an experimental decision: it
makes the repair mechanism visible by guaranteeing that every offspring
needs it, which is the configuration Task 2 and Task 3 require for the
constraint-handling comparison.

Reproducibility is enforced at the run level. Every run is identified by the
pair (configuration hash, random seed); the configuration hash is a digest of
the full parameter dictionary so that two runs with different but equivalent
parameter spellings produce the same identifier. Per-generation best fitness,
mean fitness, and Hamming diversity are written to CSV files keyed by that
pair, and the final tours are stored separately. This allows any reported
figure, table, or statistical test to be regenerated from the stored data
without re-running the GA.
