%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	TASK 3 - Evaluation Methodology and Analysis
%   Maps to brief Tasks 3.1, 3.2, 3.3.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\lhead{Task 3. \emph{Evaluation Methodology and Analysis}}
\chapter{Task 3: Evaluation Methodology and Analysis}
\label{Task3}

\section{Parameter Sensitivity Analysis (Task 3.1)}
\label{T3:Sensitivity}

We ran a full-factorial sweep on kroA100 using the repair-based pipeline
(naive single-point crossover, swap mutation, duplicate/missing-city repair
on every offspring). The swept parameters and their levels are population
size $\in \{50, 100, 200\}$, crossover rate $\in \{0.70, 0.85, 0.95\}$,
mutation rate $\in \{0.01, 0.05, 0.10\}$, and selection method
$\in \{\text{tournament}, \text{roulette}\}$. This gives
$3 \times 3 \times 3 \times 2 = 54$ unique configurations. We executed
each one for 30 independent seeds at a fixed budget of 500 generations,
producing $54 \times 30 = 1{,}620$ runs. The shared instance, fitness, and
termination settings are listed in \sref{T1:Setup}.

\tref{tab:sweep-top} reports the five best configurations ranked by mean
best fitness over 30 seeds. The gap column expresses the mean as a
percentage above the known optimal tour length of 21,282.

\begin{table}[htbp]
    \centering
    \caption{Top-5 configurations from the parameter sweep, ranked by mean best fitness (lower is better). Gap is relative to the known optimal tour length of 21,282 for kroA100.}
    \label{tab:sweep-top}
    \begin{tabular}{r r r l r r r}
        \hline
        Pop.\ size & $r_c$ & $r_m$ & Selection & Mean best & Std. & Gap \\
        \hline
        200 & 0.95 & 0.10 & tournament & 49{,}346.80 & 3{,}734.14 & 131.9\% \\
        200 & 0.70 & 0.10 & tournament & 49{,}933.91 & 3{,}414.25 & 134.6\% \\
        200 & 0.85 & 0.10 & tournament & 50{,}072.36 & 3{,}782.92 & 135.3\% \\
        200 & 0.85 & 0.05 & tournament & 54{,}333.89 & 3{,}178.22 & 155.3\% \\
        200 & 0.95 & 0.05 & tournament & 55{,}066.11 & 2{,}614.14 & 158.7\% \\
        \hline
    \end{tabular}
\end{table}

All five leading configurations use tournament selection at the largest
population size with the two highest mutation rates. The per-parameter
distributions in \fref{fig:boxplot-pop}, \fref{fig:boxplot-mutation},
and \fref{fig:boxplot-selection} are computed by collapsing the sweep
over the remaining axes, so each box aggregates all runs where that
parameter level appears.

\textbf{Selection method.} Tournament selection produces a median final
tour length of 64,894 versus 86,632 for roulette. This is the largest
single-parameter effect in the sweep. Roulette assigns selection
probability proportional to fitness magnitude, and on this instance the
spread of tour lengths is small relative to the absolute magnitudes, so the
proportional weights become nearly equal and the operator behaves close
to uniform random selection. Tournament selection imposes a deterministic best-of-$k$
rule that is invariant to absolute fitness scale, keeping selection
pressure strong throughout the run.

\textbf{Mutation rate.} The medians by mutation level are 70,464 at
$r_m = 0.10$, 76,859 at $r_m = 0.05$, and 91,121 at $r_m = 0.01$. The
repair step is destructive enough that the population converges to a
small set of similar tours within the first 100--200 generations.
Without sufficient mutation, diversity cannot be restored, and the
remaining generations contribute almost nothing. At $r_m = 0.01$ the
typical offspring is essentially identical to its parents after crossover
and repair, so the search stalls.

\textbf{Population size.} The medians are 76,355 at pop=200, 81,323 at
pop=100, and 82,998 at pop=50. This is the smallest of the three effects.
Larger populations give crossover a broader pool of building blocks, but
the additional benefit is small compared with switching selection method
or raising mutation rate.

Crossover rate $r_c$ varies across the top configurations from 0.70 to
0.95 without a clear ordering, and the sweep table reflects this:
$r_c = 0.95$ wins by a small margin but $r_c = 0.70$ ranks second, ahead
of $r_c = 0.85$. The differences between crossover rates are therefore
too small to separate from noise with only 30 seeds per cell.

We chose pop\_size~$=$~200, $r_c$~$=$~0.95, $r_m$~$=$~0.10, tournament
selection as the best configuration. This is the top row of
\tref{tab:sweep-top}.

\section{Performance Evaluation (Task 3.2)}
\label{T3:Performance}

\textbf{Solution quality.} At the chosen best configuration we get a mean
best tour length of 49,347 with a standard deviation of 3,734, a gap of
131.9\% from the known optimum of 21,282. The best single seed reaches
42,623 (gap 100.3\%). These gaps are large for a 100-city instance, but
they are consistent with the operator stack: naive single-point crossover
is not feasibility-preserving, the random-insertion repair injects no
problem-specific guidance, and no local-search component (2-opt,
Lin--Kernighan, or similar) is applied. We are searching the permutation
space using only structural recombination and random perturbation, with no
TSP-specific heuristic to refine tours.

\textbf{Convergence rate.} The convergence curves show that most of the
improvement happens in the first roughly 200 generations. After that the
mean best fitness levels off. The 500-generation budget is well beyond
this point, so each run has enough generations to finish converging.

\textbf{Stability across trials.} Standard deviations across the top
configurations range from roughly 2,600 to 3,800 tour-length units,
about 5--8\% of the mean. The coefficient of variation is moderate and
consistent across the top five rows in \tref{tab:sweep-top}, and the
boxplots show no extreme outliers at the chosen setting. Run-to-run variability is driven
by where each seed's population converges in the first 100--200
generations, not by late-stage divergence.

\textbf{Computational efficiency.} The full sweep of 1,620 runs completed
in roughly 15 minutes of wall time using 8 worker processes, with per-run
cost on the order of 0.5--1 second. The repair step adds $O(n)$ work per
offspring, which is small relative to the $O(n)$ fitness evaluation.

\section{Critical Analysis of Repair Mechanism (Task 3.3)}
\label{T3:RepairCritique}

We compared three constraint-handling strategies in a three-arm
experiment. All three arms share the same GA
hyperparameters (population size 100, crossover rate 0.8, mutation
rate 0.2, tournament selection, swap mutation, 500 generations, 30 seeds),
so the only thing that differs is how we enforce the feasibility
constraint.

The three strategies are:
\begin{itemize}
    \item \textbf{Repair} (the primary pipeline). Naive single-point crossover
    can introduce duplicate cities and missing cities. After crossover and
    mutation the offspring is repaired by random-insertion: duplicates are
    removed and missing cities are inserted at random positions. Every
    evaluated chromosome is a valid tour.
    \item \textbf{Penalty.} Naive single-point crossover with no repair.
    Infeasible chromosomes are evaluated and assigned a penalty of
    $w \cdot v$ added to fitness, where $w = 50{,}000$ is the
    penalty weight and $v$ is the number of constraint violations
    (duplicate or missing cities). Infeasible individuals survive in the
    population but face strong selection pressure against them.
    \item \textbf{PMX (feasibility-preserving).} Partially-mapped crossover
    is used in place of naive crossover. PMX is closed over the set of
    permutations by construction, so no repair or penalty is required.
\end{itemize}

\tref{tab:repair-summary} summarises the final best-fitness distribution
over 30 seeds for each strategy, and the distributional view is given in
\fref{fig:boxplot-strategy}.

\begin{table}[htbp]
    \centering
    \caption{Final best-fitness summary over 30 seeds per strategy from the three-arm comparison. All arms share pop=100, $r_c$=0.8, $r_m$=0.2, tournament selection, 500 generations. Gap is relative to the optimum 21,282.}
    \label{tab:repair-summary}
    \begin{tabular}{l r r r r r r r}
        \hline
        Strategy & Count & Mean & Std. & Min & Median & Max & Gap \\
        \hline
        Repair  & 30 & 49{,}777.84 & 3{,}186.86 & 43{,}890.51 & 49{,}406.18 & 56{,}629.20 & 133.9\% \\
        PMX     & 30 & 50{,}654.82 & 3{,}572.54 & 42{,}975.87 & 51{,}053.96 & 57{,}691.28 & 138.0\% \\
        Penalty & 30 & 57{,}947.91 & 3{,}284.51 & 48{,}091.74 & 58{,}079.43 & 64{,}585.95 & 172.3\% \\
        \hline
    \end{tabular}
\end{table}

\textbf{Effectiveness of repair.} Repair attains the lowest mean tour
length of the three arms: 49,778 versus 50,655 for PMX and 57,948 for
penalty. The repair--PMX margin is roughly 1.7\% of the mean and is
small compared with the within-arm spread (standard deviations near
3,200--3,600), so the two feasibility-respecting strategies are
practically tied on this instance. Penalty is worse than both by roughly
16--17\%, well outside the within-arm spread. The minimum single-seed
values follow the same ordering: PMX produces the best individual seed at
42,976 with repair close behind at 43,891, while penalty's best is 48,092.

\textbf{Comparison with alternatives.} Penalty is the weakest of the three,
trailing repair by about 16\% in mean tour length. The penalty weight of
50,000 is enough to disfavour infeasible individuals in tournaments, but
the GA still spends selection pressure on chromosomes that contribute no
usable tour. PMX is the more informative comparator since it removes the
feasibility problem at the operator level. Repair only beats PMX by about
2\%, which shows that the repair-based pipeline performs comparably to a
feasibility-preserving alternative on kroA100. On this instance the choice
between the two is more about design preference than performance, as the
boxplot spread in \fref{fig:boxplot-strategy} shows.
